Variants of G\"{o}del's Ontological Proof \\ in a Natural Deduction Calculus

### 2016-04-01 Papers/InProgress/NaturalDeduction/GodProof-ND.tex
Add ToDo
Anderson proof update

%Anderson modification of the axioms in A1 equivalence removed and definition of essence altered. 
%Anderson's implication instead of equivalence. 
\begin{axiom}[Emended]
\label{A1}
If the negation of a property is positive, the property is not positive:
$$
\all \varphi. [P(\neg \varphi) \imp \neg P(\varphi)]
$$
\end{axiom}

\begin{axiom}[As in Scott's]
\label{A2}
A property necessarily implied by a positive property is positive:
$$
\all \varphi. \all \psi.[(P(\varphi) \wedge \nec \all x.[\varphi(x) \imp \psi(x)]) \imp P(\psi)]
$$
\end{axiom}

\begin{theorem}
\label{T1}
Positive properties are possibly exemplified:
$$
\all \varphi. [P(\varphi) \imp \pos \ex x.\varphi(x)]
$$
\end{theorem}
\begin{proof}
This proof is omitted, because any of the proofs of Theorem 1 in Sections \ref{sec:scott} and \ref{sec:newproof} would suffice to prove this theorem with only small modifications: the equivalence of axiom \ref{A1} would not need to be reduced to an implication by conjunction elimination. 
\end{proof}

\begin{definition}[Emended]
\label{D1}
A \emph{God-like} being necessarily possesses those and only those properties which are positive:
$$
G(x) \biimp \forall \varphi. [P(\varphi) \biimp \nec \varphi(x)]
$$
\end{definition}

\begin{axiom}[As in Scott's]
\label{A3}
The property of being god-like is positive:
$$
P(G)
$$
\end{axiom}

\begin{corollary}
\label{C1}
Possibly, a God-like being exists:
$$
\pos \ex x. G(x)
$$
\end{corollary}
\begin{proof}
This proof is omitted, because it is identical to the proof of the same corollary shown in Sections \ref{sec:scott} and \ref{sec:newproof}.
\end{proof}

\begin{axiom}[As in Scott's]
\label{A4}
Positive properties are necessarily positive:
$$
\all \varphi.[P(\varphi) \to \Box \; P(\varphi)]
$$
\end{axiom}

\begin{definition}[Emended]
\label{D2}
An \emph{essence} of an individual is a property possessed by it and necessarily implying those and only those properties necessarily possessed by it:
$$
\ess{\varphi}{x} \biimp \varphi(x) \wedge \all \psi. (\nec \psi(x) \biimp \nec \all x. (\varphi(x) \imp \psi(x)))
$$
\end{definition}

\begin{definition}[As in Scott's]
\label{D3}
\emph{Necessary existence} of an individual is the necessary exemplification of all its essences:
$$
E(x) \biimp \all \varphi.[\ess{\varphi}{x} \imp \nec \ex y.\varphi(y)]
$$
\end{definition}

\begin{axiom}[As in Scott's]
\label{A5}
Necessary existence is a positive property:
$$
P(E)
$$
\end{axiom}

@@

\begin{lemma}
\label{L1}
If there is a God-like being, then there is a God-like being necessarily:
$$
\ex z. G(z) \imp \nec \ex x. G(x)
$$
\end{lemma}
\begin{proof} \hfill
\begin{prooftree}
\AXC{$ $} \RightLabel{1}
\UIC{$\ex z. G(z)$}\RightLabel{$\ex_E$}
\UIC{$[\subproof_6] \quad G(\gamma)$}
\end{prooftree}

\begin{scriptsize}
\begin{prooftree}
\AXC{$\subproof_6$} \dashedLine
\UIC{$G(\gamma)$}  \dottedLine \RightLabel{D\ref{D1}}
\UIC{$\forall \varphi. [P(\lambda x.\varphi(x)) \biimp \nec \varphi(\gamma)]$}\RightLabel{$\forall_E$}
\UIC{$P(\lambda x.\rho(x)) \biimp \nec \rho(\gamma)$}
\AXC{$ $} \RightLabel{5}
\UIC{$\nec \rho(\gamma)$} \RightLabel{$\imp_E$}
\BIC{$P(\lambda x.\rho(x))$} 
\AXC{$ $} \RightLabel{6}
\UIC{$G(y) $}\dottedLine \RightLabel{D\ref{D1}}
\UIC{$\forall \varphi. [P(\lambda x.\varphi(x)) \biimp \nec \varphi(y)]$}
\UIC{$[P(\lambda x.\rho(x)) \biimp \nec \rho(y)]$}
\BIC{$\nec \rho(y)$}\RightLabel{$\nec_E$}
\UIC{$\omega':  \rho(y)$}\RightLabel{$\imp_I^6$}
\UIC{$\omega':  G(y)\imp \rho(y)$}\RightLabel{$\forall_I$}
\UIC{$\omega':  \forall y. (G(y)\imp \rho(y))$}\RightLabel{$\nec_I$}
\UIC{$\nec \forall y. (G(y)\imp \rho(y))$}\RightLabel{$\imp_I^5$}
\UIC{$\nec \rho(\gamma)\imp \nec \forall y. (G(y)\imp \rho(y))$}
\end{prooftree}
\end{scriptsize}

\begin{scriptsize}
\begin{prooftree}
\AXC{$\nec P(\rho)\imp \nec \forall y .(G(y)\imp \rho(y))$} \RightLabel{$\imp_E$}
\AXC{$ $} \RightLabel{6}
\UIC{$\nec \rho(\gamma)$} \RightLabel{}
\UIC{$\rho(\gamma)$} \RightLabel{$\nec_E$}
\AXC{$\subproof_7$} \dashedLine
\UIC{$\omega': \rho(\gamma)\imp\Box P(\rho)$}\RightLabel{$\imp_E$}
\BIC{$\nec P(\rho)$} \RightLabel{$\imp_E$}
\BIC{$\nec \forall y .(G(y)\imp \rho(y))$}\RightLabel{$\imp_I^6$}
\UIC{$\rho(\gamma) \imp \nec \forall y.(G(y) \imp \rho(y))$}\RightLabel{$\forall_I$}
\UIC{$\forall \psi.(\psi(\gamma) \imp \nec \forall y.(G(y) \imp \psi(y)))$}
\AXC{$\subproof_6$} \dashedLine
\UIC{$G(\gamma)$}  
\BIC{$G(\gamma) \wedge \forall \psi.(\psi(\gamma) \imp \nec \forall y.(G(y) \imp \psi(y)))$}\dottedLine \RightLabel{D\ref{D2}}
\UIC{$[\subproof_8] \qquad \ess{G}{\gamma}$}
\end{prooftree}
\end{scriptsize}

%implication for equivalence in essence: 
\begin{small}
\begin{prooftree}
\AXC{$ $} \RightLabel{7}
\UIC{$\nec  \forall y. [G(y)\imp \varphi(y)]$}\RightLabel{$\nec_E$}
\UIC{$\omega'':   \forall y. [G(y)\imp \varphi(y)]$}\RightLabel{$\forall_E$}
\UIC{$\omega'':   G(\gamma)\imp \varphi(\gamma)$}
\AXC{$\subproof_6$} \dashedLine
\UIC{$\omega'':  G(\gamma)$}  \RightLabel{$\imp_E$}
\BIC{$\omega'':   \varphi(\gamma)$}\RightLabel{$\nec_I$}
\UIC{$\nec \varphi(\gamma)$}\RightLabel{$\imp_I^7$}
\UIC{$\nec  \forall y. [G(y)\imp \varphi(y)]\imp \nec \varphi(\gamma)$}
\end{prooftree}
\end{small}

\begin{small}
\begin{prooftree}
\AXC{$\subproof_8$} \dashedLine
\UIC{$\nec \varphi(\gamma)\biimp \nec  \forall y. [G(y)\imp \varphi(y)] $}
\AXC{$ $} \RightLabel{6}
\UIC{$G(x)$}\RightLabel{$\wedge_I$}
\BIC{$\ess{G}{\gamma}$}
\AXC{Axiom \ref{A5}} \dashedLine
\UIC{$P(E)$}\dottedLine 
\AXC{$\subproof_6$} \dashedLine
\UIC{$G(\gamma)$}\dottedLine \RightLabel{D\ref{D1}}
\UIC{$\forall \varphi.(P(\varphi) \biimp \nec \varphi(\gamma))$}\RightLabel{$\forall_E$}
\UIC{$P(E) \biimp \nec E(\gamma)$}\RightLabel{$\imp_E$}
\BIC{$\nec E(\gamma)$}
    \AXC{Axiom T} \dashedLine
    \UIC{$\nec E(\gamma) \imp E(\gamma)$} \RightLabel{$\imp_E$}
\BIC{$ E(\gamma)$} \dottedLine \RightLabel{D\ref{D3}}
\UIC{$ \all \varphi.[\ess{\varphi}{\gamma} \imp \nec \ex x.\varphi(x)] $}\RightLabel{$\all_E$}
\UIC{$ \ess{G}{\gamma} \imp \nec \ex x. G(x) $}\RightLabel{$\imp_E$}
\BIC{$\nec \ex x. G(x)$} \RightLabel{$\imp_I^1$}
\UIC{$\ex z. G(z) \imp \nec \ex x. G(x)$}
\end{prooftree}
\end{small}
\end{proof}

\setcounter{theorem}{2}
\begin{theorem}
\label{T3}
Necessarily, there exists a God-like being:
$$
\nec \ex x. G(x)
$$
\end{theorem}
\begin{proof}
This proof is identical to the proof of Theorem 3 shown in Section \ref{sec:newproof}.
\end{proof}

@@

ToDo: should we try to show here also an alternative proof of T3 using only Anderson's axioms A1, A2 and A3 and definition D1??  I don't know how hard this would be. Hájek1996 could be helpful.

@@

>>>>

%Anderson modification of the axioms in A1 equivalence removed and definition of essence altered. 
%Anderson's implication instead of equivalence. 
\begin{axiom}[Emended]
\label{A:A1}
If the negation of a property is positive, the property is not positive:
$$
\all \varphi. [P(\neg \varphi) \imp \neg P(\varphi)]
$$
\end{axiom}

\begin{axiom}[As in Scott's]
\label{A:A2}
A property necessarily implied by a positive property is positive:
$$
\all \varphi. \all \psi.[(P(\varphi) \wedge \nec \all x.[\varphi(x) \imp \psi(x)]) \imp P(\psi)]
$$
\end{axiom}

\begin{theorem}
\label{A:T1}
Positive properties are possibly exemplified:
$$
\all \varphi. [P(\varphi) \imp \pos \ex x.\varphi(x)]
$$
\end{theorem}
\begin{proof}
This proof is omitted, because any of the proofs of Theorem 1 in Sections \ref{sec:scott} and \ref{sec:newproof} would suffice to prove this theorem with only small modifications: the equivalence of axiom \ref{A1} would not need to be reduced to an implication by conjunction elimination. 
\end{proof}

\begin{definition}[Emended]
\label{A:D1}
A \emph{God-like} being necessarily possesses those and only those properties which are positive:
$$
G(x) \biimp \forall \varphi. [P(\varphi) \biimp \nec \varphi(x)]
$$
\end{definition}

\begin{axiom}[As in Scott's]
\label{A:A3}
The property of being god-like is positive:
$$
P(G)
$$
\end{axiom}

\begin{corollary}
\label{A:C1}
Possibly, a God-like being exists:
$$
\pos \ex x. G(x)
$$
\end{corollary}
\begin{proof}
This proof is omitted, because it is identical to the proof of the same corollary shown in Sections \ref{sec:scott} and \ref{sec:newproof}.
\end{proof}

\begin{axiom}[As in Scott's]
\label{A:A4}
Positive properties are necessarily positive:
$$
\all \varphi.[P(\varphi) \to \Box \; P(\varphi)]
$$
\end{axiom}

\begin{definition}[Emended]
\label{A:D2}
An \emph{essence} of an individual is a property possessed by it and necessarily implying those and only those properties necessarily possessed by it:
$$
\ess{\varphi}{x} \biimp \varphi(x) \wedge \all \psi. (\nec \psi(x) \biimp \nec \all x. (\varphi(x) \imp \psi(x)))
$$
\end{definition}

\begin{definition}[As in Scott's]
\label{A:D3}
\emph{Necessary existence} of an individual is the necessary exemplification of all its essences:
$$
E(x) \biimp \all \varphi.[\ess{\varphi}{x} \imp \nec \ex y.\varphi(y)]
$$
\end{definition}

\begin{axiom}[As in Scott's]
\label{A:A5}
Necessary existence is a positive property:
$$
P(E)
$$
\end{axiom}

@@

ToDo: Proof of $\star_7$ is missing in the proof of lemma 1 below.

\begin{lemma}
\label{A:L1}
If there is a God-like being, then there is a God-like being necessarily:
$$
\ex z. G(z) \imp \nec \ex x. G(x)
$$
\end{lemma}
\begin{proof} \hfill
\begin{prooftree}
\AXC{$ $} \RightLabel{1}
\UIC{$\ex z. G(z)$}\RightLabel{$\ex_E$}
\UIC{$[\subproof_6] \quad G(\gamma)$}
\end{prooftree}

\begin{scriptsize}
\begin{prooftree}
\AXC{$\subproof_6$} \dashedLine
\UIC{$G(\gamma)$}  \dottedLine \RightLabel{D\ref{A:D1}}
\UIC{$\forall \varphi. [P(\lambda x.\varphi(x)) \biimp \nec \varphi(\gamma)]$}\RightLabel{$\forall_E$}
\UIC{$P(\lambda x.\rho(x)) \biimp \nec \rho(\gamma)$}
\AXC{$ $} \RightLabel{5}
\UIC{$\nec \rho(\gamma)$} \RightLabel{$\imp_E$}
\BIC{$P(\lambda x.\rho(x))$} 
\AXC{$ $} \RightLabel{6}
\UIC{$G(y) $}\dottedLine \RightLabel{D\ref{A:D1}}
\UIC{$\forall \varphi. [P(\lambda x.\varphi(x)) \biimp \nec \varphi(y)]$}
\UIC{$[P(\lambda x.\rho(x)) \biimp \nec \rho(y)]$}
\BIC{$\nec \rho(y)$}\RightLabel{$\nec_E$}
\UIC{$\omega':  \rho(y)$}\RightLabel{$\imp_I^6$}
\UIC{$\omega':  G(y)\imp \rho(y)$}\RightLabel{$\forall_I$}
\UIC{$\omega':  \forall y. (G(y)\imp \rho(y))$}\RightLabel{$\nec_I$}
\UIC{$\nec \forall y. (G(y)\imp \rho(y))$}\RightLabel{$\imp_I^5$}
\UIC{$[\subproof_7] \quad \nec \rho(\gamma)\imp \nec \forall y. (G(y)\imp \rho(y))$}
\end{prooftree}
\end{scriptsize}

%implication for equivalence in essence: 
\begin{small}
\begin{prooftree}
\AXC{Axiom \ref{A:A3}} 
\UIC{$P(G)$}
\AXC{$ $} \RightLabel{7}
\UIC{$\nec  \forall y. [G(y)\imp \varphi(y)]$}\RightLabel{$\wedge_I$}
\BIC{$P(G)\wedge \nec  \forall y. [G(y)\imp \varphi(y)]$}\RightLabel{$\imp_E$}
\AXC{Axiom \ref{A:A2}} 
\BIC{$P(\varphi)$}
\AXC{$\subproof_6$} \dashedLine 
\UIC{$  G(\gamma)$}  \RightLabel{D\ref{A:D1}}
\UIC{$ \all \varphi. [P(\varphi)\biimp \nec \varphi(\gamma)]$}  \RightLabel{$\all_E$}
\UIC{$ P(\varphi)\biimp \nec \varphi(\gamma)$}  \RightLabel{$\biimp_E$}
\BIC{$ \nec \varphi(\gamma)$} \RightLabel{$\nec_I$}
\UIC{$[\subproof_8] \quad \nec  \forall y. [G(y)\imp \varphi(y)]\imp \nec \varphi(\gamma)$}
\end{prooftree}
\end{small}

\begin{small}
\begin{prooftree}
\AXC{$\subproof_7$ and $\subproof_8$} \dashedLine
\UIC{$\nec \varphi(\gamma)\biimp \nec  \forall y. [G(y)\imp \varphi(y)] $}
\UIC{$\forall \varphi. (\nec \varphi(\gamma)\biimp \nec  \forall y. [G(y)\imp \varphi(y)]) $}
\AXC{$ $} \RightLabel{6}
\UIC{$G(x)$}\RightLabel{$\wedge_I$}
\BIC{$\ess{G}{\gamma}$}
\AXC{Axiom \ref{A5}} \dashedLine
\UIC{$P(E)$}\dottedLine 
\AXC{$\subproof_6$} \dashedLine
\UIC{$G(\gamma)$}\dottedLine \RightLabel{D\ref{A:D1}}
\UIC{$\forall \varphi.(P(\varphi) \biimp \nec \varphi(\gamma))$}\RightLabel{$\forall_E$}
\UIC{$P(E) \biimp \nec E(\gamma)$}\RightLabel{$\imp_E$}
\BIC{$\nec E(\gamma)$}
    \AXC{Axiom T} \dashedLine
    \UIC{$\nec E(\gamma) \imp E(\gamma)$} \RightLabel{$\imp_E$}
\BIC{$ E(\gamma)$} \dottedLine \RightLabel{D\ref{A:D3}}
\UIC{$ \all \varphi.[\ess{\varphi}{\gamma} \imp \nec \ex x.\varphi(x)] $}\RightLabel{$\all_E$}
\UIC{$ \ess{G}{\gamma} \imp \nec \ex x. G(x) $}\RightLabel{$\imp_E$}
\BIC{$\nec \ex x. G(x)$} \RightLabel{$\imp_I^1$}
\UIC{$\ex z. G(z) \imp \nec \ex x. G(x)$}
\end{prooftree}
\end{small}
\end{proof}

\setcounter{theorem}{2}
\begin{theorem}
\label{A:T3}
Necessarily, there exists a God-like being:
$$
\nec \ex x. G(x)
$$
\end{theorem}
\begin{proof}
This proof is identical to the proof of Theorem 3 shown in Section \ref{sec:newproof}.
\end{proof}

@@

\subsection{A shorter proof without redundant axioms}

There is an alternative proof of theorem \ref{A:T3} using only Anderson's axioms \ref{A:A1}, \ref{A:A2} and \ref{A:A3} as well as definition \ref{A:D1}. %Hájek1996 could be helpful.
The proof requires only a modification of the proof of lemma \ref{A:L1}. The rest of the dependencies are as in the previous proof.

Theorem \ref{A:T1} is derived from axiom \ref{A:A1} and axiom \ref{A:A2}. Corollary \ref{A:C1} is derived from axiom \ref{A:A3} and theorem \ref{A:T1}. The proof of lemma \ref{A:L1} can be modified, so as to be derived from axiom \ref{A:A3} using definition \ref{A:D1}. Finally, theorem \ref{A:T3} can be derived from lemma \ref{A:L1} and corollary \ref{A:C1}.

\begin{lemma}
\label{A2:L1}
If there is a God-like being, then there is a God-like being necessarily:
$$
\ex z. G(z) \imp \nec \ex x. G(x)
$$
\end{lemma}

\begin{proof} \hfill
\begin{prooftree}
\AXC{$ $} \RightLabel{1}
\UIC{$\ex z. G(z)$}\RightLabel{$\ex_E$}
\UIC{$[\subproof_6] \quad G(\gamma)$}
\end{prooftree}

\begin{scriptsize}
\begin{prooftree}
\AXC{$\subproof_6$} \dashedLine
\UIC{$G(\gamma)$}  \dottedLine \RightLabel{D\ref{D1}}
\UIC{$\forall \varphi. [P(\lambda x.\varphi(x)) \biimp \nec \varphi(\gamma)]$}\RightLabel{$\forall_E$}
\UIC{$P(G) \biimp \nec G(\gamma)$}
\AXC{Axiom \ref{A:A3}} \RightLabel{}
\UIC{$P(G)$} \RightLabel{$\biimp_E$}
\BIC{$\nec G(\gamma)$}
\AXC{$ $} \RightLabel{5}
\UIC{$G(\gamma)$} \RightLabel{$\ex_I$}
\UIC{$\ex x.G(x)$}  \RightLabel{$\imp_I^5$}
\UIC{$G(\gamma)\imp \ex x.G(x)$} \RightLabel{$\nec_I$}
\UIC{$\nec [G(\gamma)\imp \ex x.G(x)]$}  \RightLabel{Axiom K}
\UIC{$\nec G(\gamma)\imp \nec \ex x.G(x)$} \RightLabel{$\imp_E$}
\BIC{$\nec \ex x.G(x)$}\RightLabel{$\imp_I^1$}
\UIC{$\ex z. G(z)\imp \nec \ex x.G(x)$}
\end{prooftree}
\end{scriptsize}

@@

ToDo: Fix overfull hboxes
